Type-Safe Modular Parsing

### 2016-12-20 paper/paper.tex
contributions
Some comments on S4

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present \name: a parsing combinator library that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parsing combinator approaches,
  which rely on some non-modular techniques. Fortunatelly Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.  Thus \name
  is built on top of a Packrat parsing library, but adds new parsing
  combinators to enable modular parsing. The new parsing combinators
  employ \emph{delegation-based} techniques and \emph{Object Algebras}
  to support extensibility. To evaluate \name we conduct a case study
  based on the ``Types and Programming Languages'' interpreters. The
  case study shows that \name is effective at reusing parsing code
  from existing interpreters, and the total parsing code is 60\%
  shorter than the non-modular parsing code.

>>>>

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present \name: a parsing combinator library that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parsing combinator approaches,
  which rely on some non-modular techniques. We show that Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.  \name
  builts on top of a Packrat parsing library, but adds new parsing
  combinators to enable modular parsing. The new parsing combinators
  employ \emph{delegation-based} techniques and \emph{Object Algebras}
  to support extensibility. To evaluate \name we conduct a case study
  based on the ``Types and Programming Languages'' interpreters. The
  case study shows that \name is effective at reusing parsing code
  from existing interpreters, and the total parsing code is 60\%
  shorter than the non-modular parsing code.

### 2016-12-20 paper/sections/OpenRecursion.tex
Some comments on S4

The first problem that prevents modularity of parsers built with
parser combinators is the existence of \emph{hard-coded recursive
  calls}. To solve this problem, \name employs \emph{open recursion}:
that is, recursive calls to parsers are parametrized. The resulting
programming style is similar to object-oriented programming, 
but using delegation instead of class-based inheritance.

@@

Open recursion is known as a useful feature that one method body can
``invoke another method of the same object via a special variable
called \lstinline{self} or, in some languages,
\lstinline[keywords={}]{this}'' (by Ralf Hinze). The interesting thing
is that such a variable \lstinline{self} is late-bound, or open to the
recursion, which means it can integrate some additional features
defined later, while existing methods that take \lstinline{self} as a
parameter can simply delegate their known cases to it.

For sure we cannot always leave it open; upon using it we have to
close the recursion, at which time we need to take a fix-point. Here
one might suddenly be reminded of the \lstinline{fix} function in some
functional languages like Haskell. In Haskell \lstinline{fix} is
defined as:

\begin{lstlisting}[language=Haskell,keywords={}]
fix f = let x = f x in x
\end{lstlisting}
In Scala, however, there is no library \lstinline{fix} function, but it can be easily implemented since Scala also supports lazy evaluation. See below:

\lstinputlisting[linerange=27-30]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIX

where function \lstinline{f} has a by-name parameter of type \lstinline{"=>} \lstinline{A"}.
Here we illustrate by the famous Fibonacci example. Traditionally it can be implemented in Scala using pattern-matching:
\lstinputlisting[linerange=34-40]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB

An alternative implementation using open recursion is as follows:
\lstinputlisting[linerange=44-51]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB2
where \lstinline{self} is the explicit self-reference, namely the delegator. Functions like \lstinline{fib2} which takes
an explicit self-reference as its parameter are called \textit{generators} (by Ralf Hinze).
If we close the recursion on \lstinline{fib2} only, the evaluation of \lstinline{fix(fib2)(2)}, namely the above \lstinline{x2} is processed as follows:
\begin{lstlisting}[language=Haskell,keywords={}]
   fix(fib2)(2)
= lazy val a = fib2(a); a.apply(2)
= lazy val a = fib2(a); fib2(a)(2)
= ... (the third case in fib2)
= lazy val a = fib2(a); a.apply(1) + a.apply(0)
= lazy val a = fib2(a); fib2(a)(1) + fib2(a)(0)
= ... (the first two cases in fib2)
= lazy val a = fib2(a); 1 + 0
= 1
\end{lstlisting}
which behaves similarly as \lstinline{fib(2)}. The process of evaluation will terminate if there are some base cases (like the first two in
\lstinline{fib2}) which stop the recursion in their branches, and the evaluation is finally reduced to those cases.

Suppose now we want to make use of open recursion for additional operations. We first define a combinator which combines the results of two generators in a pair:
\lstinputlisting[linerange=55-58]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_COMPOSE

@@

\subsection{Parsing with Open Recursion}\label{subsec:parsingwithopen}
Open recursion potentially opens the gate of extensibility for parsing, as a parser can now be open to additional features in the future.
Our exploration starts from a simple example, where a single parser only parses a free variable (i.e., a string). Traditionally, to write
such a parser, we need to import the library as
well as defining the datatype for which it produces.

@@

\lstinputlisting[linerange=44-49]{../Scala/Parser/src/PaperCode/ParsingWithOR.scala}% APPLY:linerange=PARSING_PLIT

@@

We proceed to implement those parsers. An issue is that we need both a \lstinline{Parser[Expr]} and a \lstinline{Parser[Type]} in the fixpoint.
Hence we use a pair in the parameter:

>>>>

The first problem that prevents modularity of parsers built with
parser combinators is the existence of \emph{hard-coded recursive
  calls}. To solve this problem, \name employs \emph{open recursion}:
that is, recursive calls to parsers are parametrized. The resulting
programming style is similar to object-oriented programming, 
but using delegation instead of class-based inheritance.
This programming style enables parsing OO-style ASTs, which 
can be naturally extended with new language constructs.

@@

Open recursion is known as a useful feature that one method body can
``invoke another method of the same object via a special variable
called \lstinline{self} or, in some languages,
\lstinline[keywords={}]{this}'' (by Ralf Hinze)\bruno{this is not how we cite references!}. The interesting thing
is that such a variable \lstinline{self} is late-bound, or open to the
recursion, which means it can integrate some additional features
defined later, while existing methods that take \lstinline{self} as a
parameter can simply delegate their known cases to it.

\paragraph{Example of Open Recursion} To illustrate the use and
usefulness of open recursion, we first consider a standard recursive 
implementation of the Fibonacci function:

\lstinputlisting[linerange=34-40]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB

\noindent An alternative implementation using open recursion is:

\lstinputlisting[linerange=44-51]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIB2
\bruno{Fix should be replaces by Open. Please coordinate with Huang to
ensure that your code and his code are coherent (i.e you use same
names and same conventions),}
\noindent This alternative implementation has an extra argument
\lstinline{self}, which serves as an explicit self-reference. In the 
definition of the function, direct recursive calls are replaced by 
calls to the self reference.

\paragraph{Closing the Open Recursion} An implementation of
 standard the standard Fibonacci function is recovered by closing 
the open recursion using a \emph{lazy fixpoint combinator}:
\bruno{definition of fib with fix here.}

In Scala the lazy fixpoint combinator is straightforwardly implemented~\cite{} as follows:

\lstinputlisting[linerange=27-30]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_FIX
\bruno{can be defined in a single line.}
\noindent Note that the function \lstinline{f} has a by-name parameter 
of type \lstinline{"=>} \lstinline{A"}. This is necessary to ensure
termination of well-founded recursive calls. In such cases, 
there are some base cases (like the first two in
\lstinline{fib2}) which stop the recursion in their branches, and the 
evaluation is finally reduced to those cases.

\begin{comment} 
If we close the recursion on \lstinline{fib2} only, the evaluation of \lstinline{fix(fib2)(2)}, namely the above \lstinline{x2} is processed as follows:
\begin{lstlisting}[language=Haskell,keywords={}]
   fix(fib2)(2)
= lazy val a = fib2(a); a.apply(2)
= lazy val a = fib2(a); fib2(a)(2)
= ... (the third case in fib2)
= lazy val a = fib2(a); a.apply(1) + a.apply(0)
= lazy val a = fib2(a); fib2(a)(1) + fib2(a)(0)
= ... (the first two cases in fib2)
= lazy val a = fib2(a); 1 + 0
= 1
\end{lstlisting}
which behaves similarly as \lstinline{fib(2)}. The process of evaluation will terminate if there are some base cases (like the first two in
\lstinline{fib2}) which stop the recursion in their branches, and the evaluation is finally reduced to those cases.
\end{comment}

\bruno{We may want to use a better example than what is next.}
Suppose now we want to make use of open recursion for additional operations. We first define a combinator which combines the results of two generators in a pair:
\lstinputlisting[linerange=55-58]{../Scala/Parser/src/PaperCode/SampleParser.scala}% APPLY:linerange=OPENRECURSION_COMPOSE

@@

\subsection{Parsing with Open Recursion}\label{subsec:parsingwithopen}

Open recursion potentially opens the gate of extensibility for
parsing, as a parser can now be open to additional features in the future.

\bruno{Most of the text below overlaps with what Huang was already
  written. I think you and Huang need to sit down, look at the text 
both of you wrote to motivate the problem, and rewrite that in the
Overview Section.}

\bruno{BEGIN: Text to be deleted/moved/rewritten in S2}
Our exploration starts from a simple example, where a single parser only parses a free variable (i.e., a string). Traditionally, to write
such a parser, we need to import the library as
well as defining the datatype for which it produces.

@@

\lstinputlisting[linerange=44-49]{../Scala/Parser/src/PaperCode/ParsingWithOR.scala}% APPLY:linerange=PARSING_PLIT
\bruno{END: Text to be deleted/moved/rewritten in S2}

\bruno{Now, here you can refer back to the example which is used in S2
to illustrate the problem. But your \emph{main goal} is to talk about
the implementation of the combinators that are helpful for Open
Recursion and parsing OO ASTs. So you need some major rewrite of the 
text that comes next. First, building on open recursion, explain the
implementation of the new parsing combinators. Then, finish of the
section with a good example of how to parse OO ASTs, perhaps
extending the example discussed in S2 a bit more.}

@@

We proceed to implement those parsers. An issue is that we need both a \lstinline{Parser[Expr]} and a \lstinline{Parser[Type]} in the fixpoint.
Hence we use a pair in the parameter:\bruno{I don't think the approach
below, adding pairs, is very modular: it requires changing your }
